---
title: "Quantitative methods for population dynamics"
description: "A two-day workshop with `R`"
site: distill::distill_website
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

**What**: Quantifying population dynamics is crucial for the conservation and management of animal and plant species. This workshop deals with the analysis and modelling of population dynamics. In this introductory workshop, we will cover population projection matrix models and population viability analyses, the estimation of demographic parameters (e.g. survival, dispersal) using capture-recapture models and the estimation of population density and abundance using capture-recapture, N-mixture and distance sampling models. Our hope is to provide you with what you need to go your own path. The event is free of charge, and video-recorded. 

**For whom**: This is a workshop for ecologists. Basic knowledge of `R` is required. See [R for beginners](https://cran.r-project.org/doc/contrib/Paradis-rdebuts_en.pdf) by E. Paradis for an introduction.

**How**: The format of the workshop will be a combination of lectures and live demonstrations in R with packages `popbio`, `Distance`, `RMark`, `AICcmodavg` and `unmarked`.

**Who**:   

+ Lecturers: [Olivier Gimenez](https://twitter.com/oaggimenez), [Aurélien Besnard](https://twitter.com/abesnardEPHE) and [Sarah Cubaynes](https://twitter.com/sarahcubaynes).  

+ Helpers: Coline Cannone, Thierry Chambert, [Thibaut Couturier](https://twitter.com/ThibautCouturie), [Valentin Lauret](https://twitter.com/ValentinLauret), [Marwan Naciri](https://twitter.com/MarwanNaciri2), [Javi Fernández-López](https://twitter.com/javi_ferlop), [Lise Viollat](https://twitter.com/LiseViollat)

**When**: Once a year

+ June 1-2, 2023 <br> Mornings 9am-12am and afternoons 2pm-5pm (Paris time UTC/GMT +2 hours)

+ ~~March 21-22, 2022 <br> Mornings 9am-12am and afternoons 2pm-5pm (Paris time UTC/GMT +2 hours)~~


**Where**: On YouTube     

+ 2022 edition: Video recordings available at: <https://youtube.com/playlist?list=PL5CiXbb9zgjSS_nJUiLKsIFKOSPINS3P6>

+ 2023 edition: Video recordings available at: <https://youtube.com/playlist?list=PL5CiXbb9zgjSP_KeorqJIBh7iH5GkC38->

### Program 

+ **Estimating abundance and density of populations**
1. Conventional methods and the detectability issue
2. Capture-recapture, distance sampling and N-mixture

+ **Estimation of demographic parameters using capture-recapture models**
1. Survival estimation from capture-recapture data
3. Using covariates to explain variation
4. Inference about dispersal

+ **Population projection models and population viability analyses**
1. Count-based projection models
2. Age- and stage-structured models, sensitivity analyses
3. Deterministic and stochastic models

### Schedule

**June 1, 2023**         
9am-9:50am - Introduction, conventional methods and the detectability issue, capture-recapture for closed populations part 1.      
10am-10:50am - Capture-recapture for closed populations part 2.     
11am-11:50am - Distance sampling.     
2pm-2:50pm - N-mixture models.     
3pm-3:50pm - Survival estimation.     
4pm-4:50pm - Survival estimation with covariates.     

**June 2, 2023**            
9am-9:50am - Capture-recapture multistate models.     
10am-10:50am - Count-based PVA.     
11am-11:50am - PVA with matrix models - determinist.     
2pm-2:50pm - PVA with matrix models - stochastics.     
3pm-3:30pm - Conclusion.     


### Requirements

+ Install [`R`](https://cloud.r-project.org/) and [`RStudio`](https://rstudio.com/products/rstudio/download/#download). `RStudio` is recommended but not mandatory. 

+ Install the R package `RMark` following the steps described at <http://www.phidot.org/software/mark/rmark/>. For Mac users, the installation is a bit tricky, check out the procedure [here](tutorials/rmark-mac.html).

+ Install the R packages `unmarked`, `Distance`, `AICcmodavg` and `popbio` directly from the console with
```{r eval = FALSE, echo = TRUE}
install.packages(c("unmarked", "popbio", "Distance", "AICcmodavg"))
```

### Reuse {.appendix}

Text and figures are licensed under Creative Commons Attribution [CC BY 4.0](https://creativecommons.org/licenses/by/4.0/). Any computer code (R, HTML, CSS, etc.) in slides and worksheets, including in slide and worksheet sources, is also licensed under [MIT](https://github.com/oliviergimenez/popdyn-workshop/blob/master/LICENSE.md). 
