---
title: "Quantitative methods for population dynamics"
description: "A two-day workshop with `R`"
site: distill::distill_website
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

**What**: Quantifying population dynamics is crucial for the conservation and management of animal and plant species. This workshop deals with the analysis and modelling of population dynamics. In this introductory workshop, we will cover population projection matrix models, the estimation of demographic parameters (e.g. survival, dispersal) using capture-recapture models and the estimation of population density and abundance using capture-recapture, N-mixture and distance sampling models. Our hope is to provide you with what you need to go your own path. The event is free of charge, and video-recorded. 

**For whom**: This is a workshop for ecologists. Basic knowledge of `R` is required. See [R for beginners](https://cran.r-project.org/doc/contrib/Paradis-rdebuts_en.pdf) by E. Paradis for an introduction.

**How**: The format of the workshop will be a combination of lectures and live demonstrations in R with packages `popbio`, `RMark` and `unmarked`.

**Who**: [Olivier Gimenez](https://twitter.com/oaggimenez), [Aurélien Besnard](https://twitter.com/abesnardEPHE) and [Sarah Cubaynes](https://twitter.com/sarahcubaynes).  

**When**: March 21-22, 2022

**Where**: Online

### Program 

+ **Population projection matrix models**
1. Introduction to matrix models
2. Age- and stage-structured models
3. Leslie models as a particular case; sensitivity analysis.
4. Introduction to `popbio`
5. Further analysis issues: density-dependence; random environment; demographic stochasticity / branching processes; multisite and multistage models

+ **Estimation of survival, transition between reproductive states, productivity using capture-recapture and generalized linear models**
1. Survival estimation from capture-recapture data
2. Multiple group, age and cohort models
3. Incorporation of external covariates
4. Introduction to `RMark`
5. Further analysis issues: estimation of recruitment and population growth rate by reversing capture histories; multi-state models allowing estimation of transitions (including movements)

+ **Estimating abundance and density of populations**
1. Introduction to counting methods 
2. Conventional methods and the detectability issue
3. Capture-recapture, N-mixture and line transect samplings
4. Introduction to `unmarked`
5. Further analysis issues: introducing covariates; heterogeneity

### Requirements

+ Install [`R`](https://cloud.r-project.org/) and [`RStudio`](https://rstudio.com/products/rstudio/download/#download). `RStudio` is recommended but not mandatory. 

+ Install the R package `Rmark` following the steps described at <http://www.phidot.org/software/mark/rmark/>.

+ Install the R packages `unmarked` and `popbio` directly from the console with
```{r eval = FALSE, echo = TRUE}
install.packages(c("unmarked", "popbio"))
```

### Reuse {.appendix}

Text and figures are licensed under Creative Commons Attribution [CC BY 4.0](https://creativecommons.org/licenses/by/4.0/). Any computer code (R, HTML, CSS, etc.) in slides and worksheets, including in slide and worksheet sources, is also licensed under [MIT](https://github.com/oliviergimenez/popdyn-workshop/blob/master/LICENSE.md). 
