---
title: "Estimation of demographic parameters - tutorial"
author: "Olivier Gimenez for the team"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9' 
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE, echo=FALSE, cache = FALSE, warning = FALSE}
options(htmltools.dir.version = FALSE,
        scipen = 1, 
        digits = 2)
knitr::opts_chunk$set(comment = "", 
                      message = FALSE, 
                      warning = FALSE)
library(tidyverse)
theme_set(theme_light())
update_geom_defaults("point", list(size = 2))
```

class: center, middle
background-size: cover

## Live demo #1: A simple CJS model

---
## Load RMark package

```{r}
library(RMark)
```

---
## Read in data (1)

Read in csv file:
```{r}
dipper_csv <- read_csv("dat/dipper.csv")
```

---
## Read in data (2)

```{r}
dipper_csv
```

---
## Convert to Mark format (1)

Build a data.frame with encounter histories, sex and (scaled) wing length:
```{r}
dipper <- data.frame(ch = unite(dipper_csv %>%
                                  select(year_1981:year_1987), 
                                col = "ch", 
                                sep = ""), 
                     sex = dipper_csv$sex,
                     wglength = scale(dipper_csv$wing_length))
```

---
## Convert to Mark format (2)

Inspect first lines:
```{r}
head(dipper)
```

---
## Fit model with parameters constant over time

```{r warning=FALSE}
phi.p <- mark(dipper)
```

---
## Explore results (1)

We get a list with many components:
```{r}
str(phi.p, max.level = 0)
```

---
## Explore results (2)

What are the components of this list?
```{r}
names(phi.p)
```

---
## Explore results (3)

Get parameters estimates:
```{r}
phi.p$results$real
```

---
class: center, middle
background-size: cover

## Variation around the CJS model

---
## Preliminaries

+ Process data (number of capture occasions, time intervals, groups):
```{r}
dipper.proc <- process.data(data = dipper, groups = "sex")
```

--

+ Create design matrix with time and group structure:
```{r}
dipper.ddl <- make.design.data(dipper.proc)
```

--

+ Define structure for survival and detection probabilities:
```{r}
phit <- list(formula=~time) # time 
phi <- list(formula=~1)     # constant
pt <- list(formula=~time)   # time
p <- list(formula=~1)       # constant
```

---
## Fit CJS model to dipper data

```{r warning=FALSE, message=FALSE, eval = FALSE}
phit.pt <- mark(dipper.proc,
                dipper.ddl,
                model.parameters = list(Phi = phit, p = pt))
```

---

```{r warning=FALSE, message=FALSE, echo = FALSE}
phit.pt <- mark(dipper.proc,
                dipper.ddl,
                model.parameters = list(Phi = phit, p = pt))
```

---

```{r}
phit.pt$results$real
```

---
## Fit model with time-varying survival and constant detection

```{r warning=FALSE, message=FALSE, eval = FALSE}
phit.p <- mark(dipper.proc,
                dipper.ddl,
                model.parameters = list(Phi = phit, p = p))
```

---

```{r warning=FALSE, message=FALSE, echo = FALSE}
phit.p <- mark(dipper.proc,
                dipper.ddl,
                model.parameters = list(Phi = phit, p = p))
```

---
## Fit model with constant survival and time-varying detection

```{r warning=FALSE, message=FALSE, eval = FALSE}
phi.pt <- mark(dipper.proc,
                dipper.ddl,
                model.parameters = list(Phi = phi, p = pt))
```

---

```{r warning=FALSE, message=FALSE, echo = FALSE}
phi.pt <- mark(dipper.proc,
                dipper.ddl,
                model.parameters = list(Phi = phi, p = pt))
```

---
## Model ranking with AIC 

```{r}
collect.models()
```

+ Best model has constant parameters

---
class: center, middle
background-size: cover

## Live demo #2: Group effect

---
## Define sex-specific survival

```{r}
phisex <- list(formula=~sex)
```

---
## Fit model with sex-specific survival, and constant detection

```{r warning=FALSE, message=FALSE, eval = FALSE}
phisex.p <- mark(dipper.proc,
                 dipper.ddl,
                 model.parameters = list(Phi = phisex, p = p))
```

---

```{r warning=FALSE, message=FALSE, echo = FALSE}
phisex.p <- mark(dipper.proc,
                 dipper.ddl,
                 model.parameters = list(Phi = phisex, p = p))
```

---
## Parameter estimates (1)

+ On the logit scale:
```{r}
phisex.p$results$beta
```

---
## Parameter estimates (2)

+ Female survival:
```{r}
logitphiF <- phisex.p$results$beta[1,1]
plogis(logitphiF)
```

--

Male survival:
```{r}
logitphiM <- logitphiF + phisex.p$results$beta[2,1]
plogis(logitphiM)
```

---
## Parameter estimates (3)

All at once:
```{r}
phisex.p$results$real
```


---
## Model ranking with AIC 

```{r}
collect.models()
```

---
## Model with additive effect of sex and time on survival and constant detection

```{r warning=FALSE, message=FALSE, eval = FALSE}
phisexpt <- list(formula=~sex+time)

phisexpt.p <- mark(dipper.proc,
                   dipper.ddl,
                   model.parameters = list(Phi = phisexpt, 
                                           p = p))
```

---

```{r echo=FALSE, message=FALSE, warning=FALSE}
phisexpt <- list(formula=~sex+time)
phisexpt.p <- mark(dipper.proc,
                   dipper.ddl,
                   model.parameters = list(Phi = phisexpt, p = p))
```

---

```{r}
phisexpt.p$results$real
```

---

```{r, echo = FALSE, fig.width = 7.5, fig.asp = 0.718, dev = "svg", message = FALSE, warning = FALSE, fig.align='center'}
df <- data.frame(year = c(1981:1986, 1981:1986),
                 survival = c(phisexpt.p$results$real[1:6,1],
                 phisexpt.p$results$real[7:12,1]),
                 sex = c(rep("female",6), rep("male", 6))
                 )
df %>%
  ggplot() + 
  aes(x = year, y = survival, color = sex) +
  geom_line()
```


---
## Model ranking with AIC 

```{r}
collect.models()
```


---
class: center, middle
background-size: cover

## Live demo #2: Continuous individual covariate


---
## Fit model with linear effect of wing length on survival 

```{r}
phiwl <- list(formula=~wglength)
```

--

```{r warning=FALSE, message=FALSE, eval = FALSE}
phiwl.p <- mark(dipper.proc,
                dipper.ddl,
                model.parameters = list(Phi = phiwl, 
                                        p = p))
```

---
## Parameter estimates (1)

```{r warning=FALSE, message=FALSE, echo = FALSE}
phiwl.p <- mark(dipper.proc,
                dipper.ddl,
                model.parameters = list(Phi = phiwl, p = p))
```

---
## Parameter estimates (2)

```{r echo=TRUE, message=FALSE, warning=FALSE}
phiwl.p$results$beta
```

---
## Graphical representation (1)

```{r eval = FALSE}
# standardize wing length
scaled_wl <- scale(dipper_csv$wing_length)
# build a grid of 100 values between min and max of wing length
grid_wl <- seq(min(scaled_wl), max(scaled_wl), length = 100)
# predict survival on logit scale
logit_predicted_survival <- phiwl.p$results$beta[1,1] + 
                    phiwl.p$results$beta[2,1] * grid_wl
# back-transform to get predicted survival
predicted_survival <- plogis(logit_predicted_survival)
# create data.frame
df <- data.frame(winglength = grid_wl,
                 survival = predicted_survival)
```

---
## Graphical representation (2)

```{r eval = FALSE}
# visualise
df %>% 
  ggplot() +
  aes(x = winglength, y = survival) + 
  geom_line()
```

---

```{r, echo = FALSE, fig.width = 7.5, fig.asp = 0.718, dev = "svg", message = FALSE, warning = FALSE, fig.align='center'}
scaled_wl <- scale(dipper_csv$wing_length)
grid_wl <- seq(min(scaled_wl), max(scaled_wl), length = 100)
logit_predicted_survival <- phiwl.p$results$beta[1,1] + phiwl.p$results$beta[2,1] * grid_wl
predicted_survival <- plogis(logit_predicted_survival)
df <- data.frame(winglength = grid_wl,
                 survival = predicted_survival)
df %>% 
  ggplot() +
  aes(x = winglength, y = survival) + 
  geom_line()  + 
  theme_light(base_size = 18)
```

---
## Quadratic relationship

```{r}
# square standardised covariate and add it to dipper data
dipper$wglengthsq <- scale(dipper_csv$wing_length)^2
# process data
dipper.proc <- process.data(data = dipper, 
                            groups = ("sex"))
# build design matrices
dipper.ddl <- make.design.data(dipper.proc)
```


---
## Fit model with quadratic effect of wing length on survival

```{r}
phiwl2 <- list(formula=~wglength + wglengthsq)
```

--

```{r warning=FALSE, message=FALSE, eval = FALSE}
phiwl2.p <- mark(dipper.proc,
                 dipper.ddl,
                 model.parameters = list(Phi = phiwl2, p = p))
```

---

```{r warning=FALSE, message=FALSE, echo = FALSE}
phiwl2.p <- mark(dipper.proc,
                 dipper.ddl,
                 model.parameters = list(Phi = phiwl2, p = p))
```

---

```{r echo=TRUE, message=FALSE, warning=FALSE}
phiwl2.p$results$beta
```


---

```{r, echo = FALSE, fig.width = 7.5, fig.asp = 0.718, dev = "svg", message = FALSE, warning = FALSE, fig.align='center'}
scaled_wl <- scale(dipper_csv$wing_length)
grid_wl <- seq(min(scaled_wl), max(scaled_wl), length = 100)
logit_predicted_survival <- phiwl2.p$results$beta[1,1] + phiwl2.p$results$beta[2,1] * grid_wl + phiwl2.p$results$beta[2,1] * grid_wl^2
predicted_survival <- plogis(logit_predicted_survival)
df <- data.frame(winglength = grid_wl,
                 survival = predicted_survival)
df %>% 
  ggplot() +
  aes(x = winglength, y = survival) + 
  geom_line() + 
  theme_light(base_size = 18)

```

---
## Model ranking with AIC

```{r}
collect.models()
```

---
class: center, middle
background-size: cover

## Live demo #3: Environmental covariate

---
## Flood effect on survival?

+ A flood took place toward the end of the second survival interval and continue into the beginning of the third survival interval

+ Question is whether survival was lower in the two flood years

---
## Explore further model structure
```{r}
head(dipper.ddl$Phi)
```

---
## Create dummy Flood variable

```{r}
dipper.ddl$Phi$Flood <- 0
head(dipper.ddl$Phi)
```

---
## Add flood effect

```{r}
dipper.ddl$Phi$Flood[dipper.ddl$Phi$time==2 | 
                       dipper.ddl$Phi$time==3] <- 1
head(dipper.ddl$Phi)
```

---
## Fit model with flood effect

```{r}
phiFlood <- list(formula=~Flood)
```

--

```{r warning=FALSE, message=FALSE, eval = FALSE}
phiFlood.p <- mark(dipper.proc,
                   dipper.ddl,
                   model.parameters = list(Phi = phiFlood, 
                                           p = p))
```

---

```{r warning=FALSE, message=FALSE, echo = FALSE}
phiFlood.p <- mark(dipper.proc,
                   dipper.ddl,
                   model.parameters = list(Phi = phiFlood, p = p))
```

---

```{r}
phiFlood.p$results$real
```

+ Clearly, survival in flood years was much lower than survival in non-flood years

---
## Model ranking with AIC

```{r}
collect.models()
```


---
class: center, middle
background-size: cover

## Live demo #4: Age in capture-recapture models

---
## Build and fit a model with age-dependent survival

```{r}
phi.age <- list(formula =~ age) # already coded in RMark
```

--

```{r eval = FALSE}
phiage.p <- mark(dipper.proc,
                 dipper.ddl,
                 model.parameters = list(Phi = phi.age, 
                                         p = p))
```

---

```{r echo = FALSE}
phiage.p <- mark(dipper.proc,
                 dipper.ddl,
                 model.parameters = list(Phi = phi.age, 
                                         p = p))
```


---
## Parameter estimates

```{r}
phiage.p$results$real
```

---
## What if we'd like only 2 age classes? 

+ Create a (0, 1+) age variable:
```{r}
dipper.ddl <- add.design.data(dipper.proc,
                              dipper.ddl, 
                              "Phi",
                              type = "age",
                              bins = c(0, 1, 7), # (0, 1+)
                              name = "ageclass", # var name
                              right = FALSE)
```

---
## Fit model with 2 age classes on survival

```{r}
phi.age2 <- list(formula=~ageclass) # age effect on survival
```

--

```{r eval = FALSE}
phiage2.p <- mark(dipper.proc,
                  dipper.ddl,
                  model.parameters = list(Phi = phi.age2, 
                                          p = p))
```

---

```{r echo = FALSE}
phiage2.p <- mark(dipper.proc,
                  dipper.ddl,
                  model.parameters = list(Phi = phi.age2, p = p))
```


---
## Parameter estimates 
```{r}
phiage2.p$results$real
```

```{r echo = FALSE, warning=FALSE, message=FALSE}
rm(list = ls(all = TRUE))
cleanup(ask = FALSE)
```

---
class: center, middle
background-size: cover

## Live demo #5: Multisite capture-recapture models

---

## Read in data (1)

Read in csv file:
```{r}
geese_csv <- read_csv2("dat/allgeese.csv")
```

---
## Read in data (2)

```{r}
geese_csv
```

---
## Convert to Mark format (1)

Build a data.frame with encounter histories, and number of individuals:
```{r}
geese <- data.frame(ch = unite(geese_csv %>% 
                                 select(year_1984:year_1989), 
                               col = "ch", 
                               sep = ""), 
                    freq = geese_csv$N)
```

---
## Convert to Mark format (2)

Inspect first lines:
```{r}
head(geese)
```

---
## Preliminaries 

```{r}
geese.proc <- process.data(data = geese,
                           model = "Multistrata")

```

--

```{r}
geese.ddl <- make.design.data(geese.proc)
```

--

```{r}
phi <- list(formula=~1) # survival constant 
phi.site <- list(formula=~stratum) # site-specific survival 
p <- list(formula=~1) # constant detection
psi.site <- list(formula=~-1+stratum:tostratum,
                 link='mlogit') # movements
```

---
## Run model with site effect on survival:

```{r eval = FALSE}
phis.psi.p <- mark(geese.proc,
                   geese.ddl,
                   model.parameters=list(S = phi.site,
                                         p = p,
                                         Psi = psi.site))
```

---

```{r echo = FALSE}
phis.psi.p <- mark(geese.proc,
                   geese.ddl,
                   model.parameters=list(S = phi.site,
                                         p = p,
                                         Psi = psi.site))
```



---

```{r}
phis.psi.p$results$real
```

---

```{r}
Psilist <- get.real(phis.psi.p, "Psi", vcv = TRUE)
Psi.values = Psilist$estimates
top.psi = TransitionMatrix(Psi.values[Psi.values$time == 1, ], 
                           vcv.real = Psilist$vcv.real)
top.psi
```

---
## Run model with state effect on survival 

```{r eval = FALSE}
phi.psi.p <- mark(geese.proc,
                  geese.ddl,
                  model.parameters=list(S = phi,
                                        p = p,
                                        Psi = psi.site))
```

---

```{r echo = FALSE}
phi.psi.p <- mark(geese.proc,
                  geese.ddl,
                  model.parameters=list(S = phi,
                                        p = p,
                                        Psi = psi.site))
```

---

```{r}
phi.psi.p$results$real
```


---

```{r}
Psilist <- get.real(phi.psi.p, "Psi", vcv = TRUE)
Psi.values = Psilist$estimates
top.psi = TransitionMatrix(Psi.values[Psi.values$time == 1, ], 
                           vcv.real = Psilist$vcv.real)
top.psi
```


---

```{r}
collect.models()
```



```{r echo = FALSE, warning=FALSE, message=FALSE}
rm(list = ls(all = TRUE))
cleanup(ask = FALSE)
```
