---
title: "Estimation of demographic parameters - tutorial"
author: "Olivier Gimenez for the team"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9' 
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE, echo=FALSE, cache = FALSE, warning = FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")
library(tidyverse)
theme_set(theme_light(base_size = 18))
update_geom_defaults("point", list(size = 2))
```

class: center, middle
background-size: cover

## Live demo #1: A simple CJS model

---
## Load RMark package

```{r}
library(RMark)
```

---
## Read in data (1)

Read in csv file:
```{r}
dipper_csv <- read_csv("dat/dipper.csv")
```

---
## Read in data (2)

```{r}
dipper_csv
```

---
## Convert to Mark format (1)

Build a data.frame with encounter histories, sex and (scaled) wing length:
```{r}
dipper <- data.frame(ch = unite(dipper_csv %>% select(year_1981:year_1987), 
                                col = "ch", 
                                sep = ""), 
                     sex = dipper_csv$sex,
                     wglength = scale(dipper_csv$wing_length))
```

---
## Convert to Mark format (2)

Inspect first lines:
```{r}
head(dipper)
```

---
## Fit model with parameters constant over time

```{r warning=FALSE}
phi.p <- mark(dipper)
```

---
## Explore results (1)

We get a list with many components:
```{r}
str(phi.p, max.level = 0)
```

---
## Explore results (2)

What are the components of this list?
```{r}
names(phi.p)
```

---
## Explore results (3)

Get parameters estimates:
```{r}
phi.p$results$real
```

---
class: center, middle
background-size: cover

## Variation around the CJS model

---

Processed data.frame:
```{r}
dipper.proc <- process.data(data = dipper, 
                            model = "CJS", 
                            groups = ("sex"))
```

---

Design matrix:
```{r}
dipper.ddl <- make.design.data(dipper.proc)
```

---

```{r}
phit <- list(formula=~time)
phi <- list(formula=~1)
pt <- list(formula=~time)
p <- list(formula=~1)
```

---

```{r warning=FALSE, message=FALSE, eval = FALSE}
phit.pt <- mark(dipper.proc,
                dipper.ddl,
                model.parameters = list(Phi = phit, p = pt))
```

---

```{r warning=FALSE, message=FALSE, echo = FALSE}
phit.pt <- mark(dipper.proc,
                dipper.ddl,
                model.parameters = list(Phi = phit, p = pt))
```

---

```{r}
phit.pt$results$real
```

---

```{r warning=FALSE, message=FALSE, eval = FALSE}
phit.p <- mark(dipper.proc,
                dipper.ddl,
                model.parameters = list(Phi = phit, p = p))
```

---

```{r warning=FALSE, message=FALSE, echo = FALSE}
phit.p <- mark(dipper.proc,
                dipper.ddl,
                model.parameters = list(Phi = phit, p = p))
```

---

```{r warning=FALSE, message=FALSE, eval = FALSE}
phi.pt <- mark(dipper.proc,
                dipper.ddl,
                model.parameters = list(Phi = phi, p = pt))
```

---

```{r warning=FALSE, message=FALSE, echo = FALSE}
phi.pt <- mark(dipper.proc,
                dipper.ddl,
                model.parameters = list(Phi = phi, p = pt))
```

---

```{r}
collect.models()
```

---
class: center, middle
background-size: cover

## Live demo #2: Group effect

---

```{r}
phisex <- list(formula=~sex)
```

---

```{r warning=FALSE, message=FALSE}
phisex.p <- mark(dipper.proc,
                dipper.ddl,
                model.parameters = list(Phi = phisex, p = p))
```

---

```{r}
phisex.p$results$beta
```

---

```{r}
logitphiF <- phisex.p$results$beta[1,1]
plogis(logitphiF)
```

---

```{r}
logitphiM <- logitphiF + phisex.p$results$beta[2,1]
plogis(logitphiM)
```

---

```{r}
phisex.p$results$real
```


---

```{r}
collect.models()
```

---

```{r}
phisexpt <- list(formula=~sex+time)
```

---

```{r warning=FALSE, message=FALSE, eval = FALSE}
phisexpt.p <- mark(dipper.proc,
                dipper.ddl,
                model.parameters = list(Phi = phisexpt, p = p))
```

---

```{r echo=FALSE, message=FALSE, warning=FALSE}
phisexpt.p <- mark(dipper.proc,
                dipper.ddl,
                model.parameters = list(Phi = phisexpt, p = p))
```


---

```{r}
collect.models()
```


---
class: center, middle
background-size: cover

## Live demo #2: Continuous individual covariate


---

```{r}
phiwl <- list(formula=~wglength)
```

---

```{r warning=FALSE, message=FALSE, eval = FALSE}
phiwl.p <- mark(dipper.proc,
                dipper.ddl,
                model.parameters = list(Phi = phiwl, p = p))
```

---

```{r warning=FALSE, message=FALSE, echo = FALSE}
phiwl.p <- mark(dipper.proc,
                dipper.ddl,
                model.parameters = list(Phi = phiwl, p = p))
```

---

```{r echo=FALSE, message=FALSE, warning=FALSE}
phiwl.p$results$beta
```

---

```{r eval = FALSE}
scaled_wl <- scale(dipper_csv$wing_length)
grid_wl <- seq(min(scaled_wl), max(scaled_wl), length = 100)
logit_predicted_survival <- phiwl.p$results$beta[1,1] + phiwl.p$results$beta[2,1] * grid_wl
predicted_survival <- plogis(logit_predicted_survival)
df <- data.frame(winglength = grid_wl,
                 survival = predicted_survival)
df %>% 
  ggplot() +
  aes(x = winglength, y = survival) + 
  geom_line()
```

---

```{r, echo = FALSE, fig.width = 7.5, fig.asp = 0.718, dev = "svg", message = FALSE, warning = FALSE, fig.align='center'}
scaled_wl <- scale(dipper_csv$wing_length)
grid_wl <- seq(min(scaled_wl), max(scaled_wl), length = 100)
logit_predicted_survival <- phiwl.p$results$beta[1,1] + phiwl.p$results$beta[2,1] * grid_wl
predicted_survival <- plogis(logit_predicted_survival)
df <- data.frame(winglength = grid_wl,
                 survival = predicted_survival)
df %>% 
  ggplot() +
  aes(x = winglength, y = survival) + 
  geom_line()
```

---

```{r}
dipper$wglengthsq <- scale(dipper_csv$wing_length)^2
dipper.proc <- process.data(data = dipper, 
                            model = "CJS", 
                            groups = ("sex"))
dipper.ddl <- make.design.data(dipper.proc)
```


---

```{r}
phiwl2 <- list(formula=~wglength + wglengthsq)
```

---

```{r warning=FALSE, message=FALSE, eval = FALSE}
phiwl2.p <- mark(dipper.proc,
                 dipper.ddl,
                 model.parameters = list(Phi = phiwl2, p = p))
```

---

```{r warning=FALSE, message=FALSE, echo = FALSE}
phiwl2.p <- mark(dipper.proc,
                 dipper.ddl,
                 model.parameters = list(Phi = phiwl2, p = p))
```

---

```{r echo=FALSE, message=FALSE, warning=FALSE}
phiwl2.p$results$beta
```

---

```{r eval = FALSE}
scaled_wl <- scale(dipper_csv$wing_length)
grid_wl <- seq(min(scaled_wl), max(scaled_wl), length = 100)
logit_predicted_survival <- phiwl2.p$results$beta[1,1] + phiwl2.p$results$beta[2,1] * grid_wl + phiwl2.p$results$beta[2,1] * grid_wl^2
predicted_survival <- plogis(logit_predicted_survival)
df <- data.frame(winglength = grid_wl,
                 survival = predicted_survival)
df %>% 
  ggplot() +
  aes(x = winglength, y = survival) + 
  geom_line()
```

---

```{r, echo = FALSE, fig.width = 7.5, fig.asp = 0.718, dev = "svg", message = FALSE, warning = FALSE, fig.align='center'}
scaled_wl <- scale(dipper_csv$wing_length)
grid_wl <- seq(min(scaled_wl), max(scaled_wl), length = 100)
logit_predicted_survival <- phiwl2.p$results$beta[1,1] + phiwl2.p$results$beta[2,1] * grid_wl + phiwl2.p$results$beta[2,1] * grid_wl^2
predicted_survival <- plogis(logit_predicted_survival)
df <- data.frame(winglength = grid_wl,
                 survival = predicted_survival)
df %>% 
  ggplot() +
  aes(x = winglength, y = survival) + 
  geom_line()
```

---

```{r}
collect.models()
```

---
class: center, middle
background-size: cover

## Live demo #3: Environmental covariate

---

```{r}
head(dipper.ddl$Phi)
```

---

```{r}
dipper.ddl$Phi$Flood <- 0
head(dipper.ddl$Phi)
```

---

```{r}
dipper.ddl$Phi$Flood[dipper.ddl$Phi$time==2 | dipper.ddl$Phi$time==3] <- 1
head(dipper.ddl$Phi)
```

---

```{r}
phiFlood <- list(formula=~Flood)
```

---

```{r warning=FALSE, message=FALSE, eval = FALSE}
phiFlood.p <- mark(dipper.proc,
                   dipper.ddl,
                   model.parameters = list(Phi = phiFlood, p = p))
```

---

```{r}
collect.models()
```


---
class: center, middle
background-size: cover

## Live demo #4: Age in capture-recapture models

---

```{r}
phi.age <- list(formula =~ age)
```

---

```{r}
phiage.p <- mark(dipper.proc,
                      dipper.ddl,
                      model.parameters = list(Phi = phi.age, 
                                              p = p))
```

---

```{r}
phiage.p$results$real
```

---

Create 0, 1+ age variable
```{r}
dipper.ddl <- add.design.data(dipper.proc,
                              dipper.ddl, # add 2 age-class structure to design matrix
                              "Phi",
                              type = "age",
                              bins = c(0, 1, 7),
                              name = "ageclass",
                              right = FALSE)
```

---

```{r}
phi.age2 <- list(formula=~ageclass) # age effect on survival
```

---

```{r}
phiage2.p <- mark(dipper.proc,
                  dipper.ddl,
                  model.parameters = list(Phi = phi.age2, p = p))
```

---

```{r}
phiage2.p$results$real
```


---
class: center, middle
background-size: cover

## Live demo #5: Multisite capture-recapture models

```{r echo = FALSE, warning=FALSE, message=FALSE}
rm(list = ls(all = TRUE))
cleanup(ask = FALSE)
```
